Ford Fusion

### 2024-09-12 mss.tex
Post

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              Manuscript submitted for review.\\
              % {\it Urbit Systems Technical Journal} I:1 (2024):  1–3. \\
              ~ \\
              Address author correspondence to \authorpatp.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{Ford Fusion}
\fancyfoot[LE,RO]{\thepage}

@@

% We will adjust page numbering in final editing.
\pagenumbering{arabic}
\setcounter{page}{1}

@@

Before Ford Fusion, each commit to the Clay filesystem validated its files using filetypes (called ``marks'') defined by files in the previous commit.\footnote{See \patp{lagrev-nocfep}, pp. TODO, this issue.} This could cause bugs if the filetype definitions had changed in a backward-incompatible manner. It also meant a commit could not add both a new filetype and new files of that type; instead, you needed two commits: one to define the filetypes, and a second to add files of that new type. More theoretically, it caused history-dependence. The validated contents of files in a commit could vary based on the history of commits that led to this one.

Another way source code failed to be self-contained was that it had access to symmetry-breaking information at build-time, namely ship, desk (Urbit's answer to a git branch), and (faked) date. A build recipe should be able to shared, cached, and rerun without dependence on local conditions, so user code now no longer learns its ship, desk, or date until runtime.

Source code also had build-time access to Urbit's immutable global namespace, called the “scry namespace”, which the kernel makes available as an implicit argument to userspace Hoon code. The scry namespace is immutable and referentially transparent, i.e. a request must always yield the same result for all time, but if an agent asks the kernel for a resource that’s from the future, hosted on another ship, or to which that agent doesn’t have permission, the kernel will deny the request.

@@

An update to one layer necessitates a reload of all layers above it; e.g. a change to Zuse should trigger updates to the vanes and userspace. Conversely, an update to a higher layer should not cause a spurious reload of lower layers, which should not be affected by the change; for example, an update to just userspace should not cause any reloads of system code.

@@

If there's a change to files in \lstinline[style=inlinecode]{/sys} on the \lstinline[style=inlinecode]{%base} desk, Clay asks Arvo to update kernelspace. Clay sends a sequence of moves (effects) to Arvo to ask Arvo to perform any necessary updates to hoon, the kernel, Zuse, and vanes. This sequence is terminated by an extra move back to Clay itself, which will be received by the updated version of Clay after migrating its state. The rebuilt Clay can then use the newly rebuilt version of Zuse to rebuild userspace and notify clients of the update. One client is Gall, which molts when Clay notifies it.

@@

A Clay commit, like a Git commit, is specified as the current value of all its changed files (and, separately, references its parent commits by hash), not as the diff from a parent commit. Unlike Git, Clay is typed, and every file must be validated according to its "mark". A mark is named like a file extension, e.g. \lstinline[style=inlinecode]{%txt}, \lstinline[style=inlinecode]{%png}, or \lstinline[style=inlinecode]{%noun}, and Clay maintains a mapping from that name to behaviors of values of that type under various operations. The last segment of any Clay path specifies the mark to use for operations on that file, including validation.

@@

Since building a file is a pure function, Clay memoizes the results of all builds, including builds of marks, mark conversions, and hoon source files. These memoization results are stored along with the desk and are used by later revisions of that desk. Future work should allow merge commits to pull memoized builds from all parents, but for now only the previous revision of the current desk is used. This is a major simplification of previous Ford architectures, which maintained much more complex caches with less clear eviction semantics. Now on every commit, we just throw away any unused memoized builds from the previous revision's Ford cache.

@@

Clay exposes file builds into the scry namespace with \lstinline[style=inlinecode]{%ca}: as an example, \lstinline[style=inlinecode]{.^(vase %ca /~zod/home/3/lib/sole/hoon)} will build the \lstinline[style=inlinecode]{sole} library.

@@

Clay exposes mark builds into the scry namespace with \lstinline[style=inlinecode]{%cb}: as an example, \lstinline[style=inlinecode]{.^(dais:clay %cb /~zod/home/3/mar/foo/hoon)} builds a \lstinline[style=inlinecode]{$dais} for the \lstinline[style=inlinecode]{%foo} mark.

@@

Clay exposes cast builds into the scry namespace with \lstinline[style=inlinecode]{%cc}: as an example, \lstinline[style=inlinecode]{.^(tube:clay %cc /~zod/home/3/foo/bar)} builds a \lstinline[style=inlinecode]{$tube} conversion gate from \lstinline[style=inlinecode]{%foo} to \lstinline[style=inlinecode]{%bar}.

@@

The \lstinline[style=inlinecode]{/=} rune imports the result of building a hoon file from a user-specified path (the second argument), wrapping it in a face specified by the first argument. The final \lstinline[style=inlinecode]{/hoon} at the end of the path must be omitted. This is mostly useful for importing a file for testing. The file at the specified path will be built as a normal userspace hoon file; i.e. its compilation subject will be Zuse augmented with the results of any Ford runes it has at the top of the file.

@@

The \lstinline[style=inlinecode]{/*} rune imports the contents of a file in the desk, specified as the third argument with the full path including the trailing mark, converted to the mark specified by the second argument, and pinned into the compilation subject wrapped in the face specified by the first argument. This can be used to import static data at build-time, such as a data file, a media file, or, in the case of this example, a hoon file as source text rather than already built.

A valid userspace hoon file must contain a nonempty list of hoons (hoon source expressions) below the Ford runes, separated by gap (more than one space, or at least one newline). The system wraps this list of hoons in a \lstinline[style=inlinecode]{=~} expression so that the result of the previous hoon is used as the subject of the next hoon. The result of the Ford runes is used as the compilation subject for this \lstinline[style=inlinecode]{=~} hoon; informally, the shape of the compilation subject can be thought of as:

@@

The \lstinline[style=inlinecode]{/~} rune imports, builds, evaluates, and pins the results of many \lstinline[style=inlinecode]{hoon} files in a directory.  Each hoon file in the specified directory will be built and evaluated. The result of evaluating each file will be added to a \lstinline[style=inlinecode]{++map} and pinned with the specified face \lstinline[style=inlinecode]{some-face}. The keys of the map will be the name of each file, and the values of the map will be the result of evaluating each file and casting its result to the type specified \lstinline[style=inlinecode]{some-type}.

>>>>

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              % Manuscript submitted for review.\\
              {\it Urbit Systems Technical Journal} I:2 (2024):  87–102. \\
              ~ \\
              Address author correspondence to \authorpatp.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{Ford Fusion}
\fancyfoot[LE,RO]{\thepage}

@@

% We will adjust page numbering in final editing.
\pagenumbering{arabic}
\setcounter{page}{87}

@@

Before Ford Fusion, each commit to the Clay filesystem validated its files using filetypes (called ``marks'') defined by files in the previous commit (see \citet{Davis2024}, pp. 33–48 in this issue). This could cause bugs if the filetype definitions had changed in a backward-incompatible manner. It also meant a commit could not add both a new filetype and new files of that type; instead, you needed two commits: one to define the filetypes, and a second to add files of that new type. More theoretically, it caused history-dependence. The validated contents of files in a commit could vary based on the history of commits that led to this one.

Another way source code failed to be self-contained was that it had access to symmetry-breaking information at build-time, namely ship, desk (Urbit's answer to a \textsc{git} branch), and (faked) date. A build recipe should be able to shared, cached, and rerun without dependence on local conditions, so user code now no longer learns its ship, desk, or date until runtime.

Source code also had build-time access to Urbit's immutable global namespace, called the “scry namespace”, which the kernel makes available as an implicit argument to userspace Hoon code. The scry namespace is immutable and referentially transparent, i.e., a request must always yield the same result for all time, but if an agent asks the kernel for a resource that’s from the future, hosted on another ship, or to which that agent doesn’t have permission, the kernel will deny the request.

@@

An update to one layer necessitates a reload of all layers above it; e.g., a change to Zuse should trigger updates to the vanes and userspace. Conversely, an update to a higher layer should not cause a spurious reload of lower layers, which should not be affected by the change; for example, an update to just userspace should not cause any reloads of system code.

@@

If there's a change to files in \lstinline[style=inlinecode]{/sys} on the \lstinline[style=inlinecode]{%base} desk, Clay asks Arvo to update kernelspace. Clay sends a sequence of moves (effects) to Arvo to ask Arvo to perform any necessary updates to Hoon, the kernel, Zuse, and vanes. This sequence is terminated by an extra move back to Clay itself, which will be received by the updated version of Clay after migrating its state. The rebuilt Clay can then use the newly rebuilt version of Zuse to rebuild userspace and notify clients of the update. One client is Gall, which molts when Clay notifies it.

@@

A Clay commit, like a Git commit, is specified as the current value of all its changed files (and, separately, references its parent commits by hash), not as the diff from a parent commit. Unlike Git, Clay is typed, and every file must be validated according to its "mark". A mark is named like a file extension, e.g., \lstinline[style=inlinecode]{%txt}, \lstinline[style=inlinecode]{%png}, or \lstinline[style=inlinecode]{%noun}, and Clay maintains a mapping from that name to behaviors of values of that type under various operations. The last segment of any Clay path specifies the mark to use for operations on that file, including validation.

@@

Since building a file is a pure function, Clay memoizes the results of all builds, including builds of marks, mark conversions, and Hoon source files. These memoization results are stored along with the desk and are used by later revisions of that desk. Future work should allow merge commits to pull memoized builds from all parents, but for now only the previous revision of the current desk is used. This is a major simplification of previous Ford architectures, which maintained much more complex caches with less clear eviction semantics. Now on every commit, we just throw away any unused memoized builds from the previous revision's Ford cache.

@@

\sloppy
Clay exposes file builds into the scry \mbox{namespace} with \lstinline[style=inlinecode]{%ca}: as an example, \lstinline[style=inlinecode]{.^(vase %ca /~zod/base/3} \lstinline[style=inlinecode]{/lib/sole/hoon)} will build the \lstinline[style=inlinecode]{sole} library.

@@

\sloppy
Clay exposes mark builds into the scry \mbox{namespace} with \lstinline[style=inlinecode]{%cb}: as an example, \lstinline[style=inlinecode]{.^(dais:clay %cb /~zod/base/3} \lstinline[style=inlinecode]{/mar/foo/hoon)} builds a \lstinline[style=inlinecode]{$dais} for the \lstinline[style=inlinecode]{%foo} mark.

@@

\sloppy
Clay exposes cast builds into the scry namespace with \lstinline[style=inlinecode]{%cc}: as an example, \lstinline[style=inlinecode]{.^(tube:clay %cc /~zod/base/3} \lstinline[style=inlinecode]{/foo/bar)} builds a \lstinline[style=inlinecode]{$tube} conversion gate from \lstinline[style=inlinecode]{%foo} to \lstinline[style=inlinecode]{%bar}.

@@

The \lstinline[style=inlinecode]{/=} rune imports the result of building a Hoon file from a user-specified path (the second argument), wrapping it in a face specified by the first argument. The final \lstinline[style=inlinecode]{/hoon} at the end of the path must be omitted. This is mostly useful for importing a file for testing. The file at the specified path will be built as a normal userspace Hoon file; i.e., its compilation subject will be Zuse augmented with the results of any Ford runes it has at the top of the file.

@@

The \lstinline[style=inlinecode]{/*} rune imports the contents of a file in the desk, specified as the third argument with the full path including the trailing mark, converted to the mark specified by the second argument, and pinned into the compilation subject wrapped in the face specified by the first argument. This can be used to import static data at build-time, such as a data file, a media file, or, in the case of this example, a Hoon file as source text rather than already built.

A valid userspace Hoon file must contain a nonempty list of \lstinline[style=inlinecode]{hoon}s (Hoon source expressions) below the Ford runes, separated by gap (more than one space, or at least one newline). The system wraps this list of \lstinline[style=inlinecode]{hoon}s in a \lstinline[style=inlinecode]{=~} expression so that the result of the previous \lstinline[style=inlinecode]{hoon} is used as the subject of the next \lstinline[style=inlinecode]{hoon}. The result of the Ford runes is used as the compilation subject for this \lstinline[style=inlinecode]{=~} \lstinline[style=inlinecode]{hoon}; informally, the shape of the compilation subject can be thought of as:

@@

The \lstinline[style=inlinecode]{/~} rune imports, builds, evaluates, and pins the results of many \lstinline[style=inlinecode]{hoon} files in a directory.  Each Hoon file in the specified directory will be built and evaluated. The result of evaluating each file will be added to a \lstinline[style=inlinecode]{++map} and pinned with the specified face \lstinline[style=inlinecode]{some-face}. The keys of the map will be the name of each file, and the values of the map will be the result of evaluating each file and casting its result to the type specified \lstinline[style=inlinecode]{some-type}.
